---
title: "LIFE748 - ML and data analysis"
author: "Francesco Del Carratore"
format: html
editor: visual
---

## Introduction

In this tutorial, we will analyze a simulated RNA-seq dataset. Our goal is to identify Differentially Expressed Genes (DEGs) between two conditions (Condition A and Condition B), visualize the results, and use both unsupervised and supervised machine learning techniques to classify the samples based on their gene expression profiles.

## Step 0: Load the required libraries

```{r warning=FALSE, message=FALSE }
library(DESeq2) # Bioconductor
library(ggplot2) # CRAN
library(EnhancedVolcano) # Bioconductor
library(pheatmap) # Bioconductor
library(caret) # CRAN
library(MASS) # CRAN
library(e1071) # CRAN
library(reshape2)

```

If these packages are not already available, you will need to install them either from CRAN (install.packages('package_name')), or from Bioconductor (BiocManager::install('package_name')

## Step 1: Load the Count Matrix

First, we load the raw count matrix and set up the sample metadata (phenotype data). We assume the first 25 columns belong to Condition A, and the next 25 belong to Condition B.

```{r}
counts <- read.csv("dataset_LIFE748_1.csv", row.names = 1)

colData <- data.frame(
  row.names = colnames(counts),
  Condition = c(rep('CondA',25),rep('CondB',25))
)

```

## Step 2: Normalize and Identify DEGs with DESeq2

We will use `DESeq2` to normalize the library sizes and perform statistical testing to identify which genes are differentially expressed between Condition B and Condition A.

```{r}
# 1. Create the DESeq2 dataset object
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = colData,
                              design = ~ Condition)

# 2. Run the DESeq2 pipeline
dds <- DESeq(dds)

# 3. Extract the results (Condition B vs Condition A)
res <- results(dds)

# 4. Convert this object into a dataframe for easier visualisation (optional)

res <- as.data.frame(res)

# 5. Extract normalized counts for downstream machine learning
vsd <- vst(dds, blind = FALSE) # Variance stabilizing transformation
normalized_counts <- assay(vsd)

```

## Step 3: Report DEGs (Table and Volcano Plot)

We define statistically significant DEGs as those with an adjusted p-value (FDR) \< 0.05 and an absolute log2 Fold Change \> 1.

```{r}
## Add gene identifiers as a column
res$gene <- rownames(res)

## Remove genes for which DESeq2 could not calculate an adjusted p-value
res_complete <- res[!is.na(res$padj), ]

## Identify significant differentially expressed genes
## Threshold: adjusted p-value < 0.05 and absolute log2 fold change > 1
DEGs <- subset(
  res_complete,
  padj < 0.05 & abs(log2FoldChange) > 1
)

## Classify direction of differential expression
DEGs$regulation <- ifelse(
  DEGs$log2FoldChange > 1,
  "Upregulated",
  "Downregulated"
)

## Number of significant DEGs
nrow(DEGs)

## Numbers up- and down-regulated
table(DEGs$regulation)

## Sort by adjusted p-value
DEGs <- DEGs[order(DEGs$padj), ]

# Display the top 20 DEGs
head(DEGs, 20)


## Export for downstream# structural bioinformatics analysis in Part 3
write.csv(
  DEGs,
  "DEG_candidates.csv",
  row.names = FALSE
)
```

Generate a volcano plot to visualise the results

```{r}
#| fig-width: 11
#| fig-height: 7
#| out-width: "100%"

EnhancedVolcano(
  res,
  lab = rownames(res),
  x = "log2FoldChange",
  y = "padj",
  pCutoff = 0.05,
  FCcutoff = 1,

  title = "Differential expression: CondB vs CondA",
  subtitle = "FDR < 0.05 and |log2FC| > 1",

  xlab = expression(Log[2] ~ "fold change"),
  ylab = expression(-Log[10] ~ "adjusted p-value"),

  selectLab = rownames(DEGs)[1:10],

  labSize = 3.5,
  pointSize = 2,

  titleLabSize = 22,
  subtitleLabSize = 12,

  legendLabels = c(
    "NS",
    expression(Log[2]~FC),
    "FDR",
    expression(FDR~and~Log[2]~FC)
  ),

  legendPosition = "right",
  legendLabSize = 10,
  legendIconSize = 3,

  drawConnectors = TRUE,
  widthConnectors = 0.5
)
```

## Step 4: Isolate Differentially Expressed Genes

To reduce noise and dimensionality for our machine learning algorithms, we will subset our normalized count matrix to only include the statistically significant DEGs identified above.

In order to make sure we don't lose too much information, let's relax a bit the thresholds for significance

```{r}
## Relax significance thresholds for machine-learning feature selection while retaining a minimum effect size requirement
ML_DEGs <- subset(
  res_complete,
  padj < 0.10 & abs(log2FoldChange) > 0.5
)

## Number of genes retained for ML
nrow(ML_DEGs)

## Extract selected genes 
ML_expression <- normalized_counts[
  rownames(normalized_counts) %in% rownames(ML_DEGs),
]

## Check
dim(ML_expression)

## Transpose
ML_data <- as.data.frame(t(ML_expression))

## Add experimental condition
ML_data$Condition <- colData$Condition

## Check
dim(ML_data)
table(ML_data$Condition)
```

## Step 5: Unsupervised Clustering

We will use unsupervised clustering to investigate whether samples group according to their gene-expression profiles without providing the algorithms with the known experimental condition labels.

### Step 5.1: k-means

```{r}
## Expression features only
X <- ML_data[, !names(ML_data) %in% "Condition"]

## Check dimensions of the clustering dataset
dim(X)

## Check for missing values
sum(is.na(X))



## Elbow test to explore the appropriate number of clusters

set.seed(748)

k.max <- 10

wss <- sapply(
  1:k.max,
  function(k) {
    kmeans(
      X,
      centers = k,
      nstart = 5,
      iter.max = 15
    )$tot.withinss
  }
)

plot(
  1:k.max,
  wss,
  type = "b",
  pch = 19,
  frame = FALSE,
  xlab = "Number of clusters (k)",
  ylab = "Total within-cluster sum of squares",
  main = "Elbow plot for k-means clustering"
)


## Perform k-means clustering using two clusters

set.seed(748)

kmeans_fit <- kmeans(
  X,
  centers = 2,
  iter.max = 15,
  nstart = 5
)

## Examine the result
kmeans_fit
```

### Step 5.2: Hierarchical Clustering

```{r}
## Calculate pairwise Euclidean distances between samples

dist_mat <- dist(
  X,
  method = "euclidean"
)

## Inspect part of the distance matrix
as.matrix(dist_mat)[1:5, 1:5]



## Perform hierarchical clustering using average linkage

hc <- hclust(
  dist_mat,
  method = "average"
)

## Plot dendrogram
plot(
  hc,
  labels = ML_data$Condition,
  main = "Hierarchical clustering of gene-expression profiles",
  xlab = "",
  sub = ""
)


## Cut the dendrogram into two clusters

hc_clusters <- cutree(
  hc,
  k = 2
)

## Compare hierarchical clusters with experimental conditions

hc_comparison <- table(
  Cluster = hc_clusters,
  Condition = ML_data$Condition
)

hc_comparison
```

### Step 5.3: Visualisation of Expression Patterns

Visualise the results (HINT: an heatmap would be perfect for this).

```{r}
#| fig-width: 10
#| fig-height: 7
#| out-width: "100%"

## Sample-condition annotation
annotation_col <- data.frame(
  Condition = ML_data$Condition
)

rownames(annotation_col) <- rownames(ML_data)

## Heatmap colour scale
heatmap_breaks <- seq(
  -3,
  3,
  length.out = 101
)

pheatmap(
  ML_expression,
  scale = "row",

  annotation_col = annotation_col,

  show_rownames = FALSE,
  show_colnames = FALSE,

  clustering_method = "average",

  border_color = NA,

  breaks = heatmap_breaks,

  main = "Expression profiles of ML-selected genes",

  fontsize = 18
)
```

## Step 6: Supervised Learning (LDA)

Next, we train classifiers to predict the condition of a sample based on its DEG expression profile. We will split the data into a training set (70%) and a testing set (30%).

```{r}
## condition to a factor
ML_data$Condition <- factor(
  ML_data$Condition,
  levels = c("CondA", "CondB")
)

## Create a reproducible 70/30 training-validation split
set.seed(748)

train_index <- createDataPartition(
  ML_data$Condition,
  p = 0.7,
  list = FALSE,
  times = 1
)

train_data <- ML_data[train_index, ]
validate_data <- ML_data[-train_index, ]

## Check
dim(train_data)
dim(validate_data)

table(train_data$Condition)
table(validate_data$Condition)
```

### Step 6.1: Train LDA (Linear Discriminant Analysis)

```{r}
## Train the LDA classifier
lda_model <- lda(
  Condition ~ .,
  data = train_data
)

## Examine
lda_model
```

### Step 6.2: Predict the Validation Samples

```{r}
lda_pred <- predict(
  lda_model,
  newdata = validate_data,
  type = "class"
)

lda_pred$class
```

### Step 6.3: LDA confusion matrix

```{r}
lda_cm <- confusionMatrix(
  data = as.factor(lda_pred$class),
  reference = as.factor(validate_data$Condition),
  positive = "CondB"
)

lda_cm

```

## Step 7: Supervised Learning (SVM)

### Step 7.1: Train the SVM classifier

```{r}
## Train an SVM classifier using the default radial kernel
svm_model <- svm(
  Condition ~ .,
  data = train_data
)

# Examine
svm_model

```

### Step 7.2: Predict the Validation Samples

```{r}
svm_pred <- predict(
  svm_model,
  newdata = validate_data
)

svm_pred
```

### Step 7.3: SVM Confusion Matrix

```{r}
svm_cm <- confusionMatrix(
  data = as.factor(svm_pred),
  reference = as.factor(validate_data$Condition),
  positive = "CondB"
)

svm_cm

```

### Step 7.4: Compare SVM Kernel Functions

```{r}
svm_radial <- svm(
  Condition ~ .,
  data = train_data,
  kernel = "radial"
)

svm_linear <- svm(
  Condition ~ .,
  data = train_data,
  kernel = "linear"
)

svm_polynomial <- svm(
  Condition ~ .,
  data = train_data,
  kernel = "polynomial"
)

svm_sigmoid <- svm(
  Condition ~ .,
  data = train_data,
  kernel = "sigmoid"
)

## Generate predictions for the validation dataset
pred_radial <- predict(svm_radial, newdata = validate_data)
pred_linear <- predict(svm_linear, newdata = validate_data)
pred_polynomial <- predict(svm_polynomial, newdata = validate_data)
pred_sigmoid <- predict(svm_sigmoid, newdata = validate_data)

```

```{r}
## Calculate confusion matrices for each SVM kernel

cm_radial <- confusionMatrix(
  pred_radial,
  validate_data$Condition,
  positive = "CondB"
)

cm_linear <- confusionMatrix(
  pred_linear,
  validate_data$Condition,
  positive = "CondB"
)

cm_polynomial <- confusionMatrix(
  pred_polynomial,
  validate_data$Condition,
  positive = "CondB"
)

cm_sigmoid <- confusionMatrix(
  pred_sigmoid,
  validate_data$Condition,
  positive = "CondB"
)


## Compare validation accuracy

svm_kernel_accuracy <- data.frame(
  Kernel = c("Radial", "Linear", "Polynomial", "Sigmoid"),
  Accuracy = c(
    cm_radial$overall["Accuracy"],
    cm_linear$overall["Accuracy"],
    cm_polynomial$overall["Accuracy"],
    cm_sigmoid$overall["Accuracy"]
  )
)

svm_kernel_accuracy
```

## Step 8: Supervised Learning (Logistic regression) 

### Step 8.1: Train the Logistic Regression Classifier

```{r}
log_model <- glm(
  Condition ~ .,
  data = train_data,
  family = "binomial"
)

##examine
summary(log_model)
```

### Step 8.2: Predict the Validation Samples

```{r}
## Predict the probability of each valildation sample belonging to CondB
log_prob <- predict(
  log_model,
  newdata = validate_data,
  type = "response"
)

## View predicted probabilities
log_prob

## Convert probabilities into class predictions

log_pred <- ifelse(
  log_prob >= 0.5,
  "CondB",
  "CondA"
)

## Convert to a factor
log_pred <- factor(
  log_pred,
  levels = c("CondA", "CondB")
)

log_pred
```

### Step 8.3: Logistic Regression Confusion Matrix

```{r}
##Compare with known experimental conditions
log_cm <- confusionMatrix(
  data = log_pred,
  reference = validate_data$Condition,
  positive = "CondB"
)

log_cm
```

## Step 9: Cross-Validation and Classifier Comparison

### Step 9.1: Create Cross-Validation Folds

```{r}
set.seed(748)

folds <- createFolds(
  ML_data$Condition,
  k = 7,
  list = TRUE,
  returnTrain = FALSE
)

## Examine the number of samples in each
sapply(folds, length)
```

### Step 9.2: LDA 7-Fold Cross-Validation

```{r}
## Create an empty vector - stores LDA predictions
lda_cv_pred <- rep(NA, nrow(ML_data))

## Perform 7-fold cross-validation
for (i in 1:length(folds)) {

  test_index <- folds[[i]]

  train_cv <- ML_data[-test_index, ]
  test_cv <- ML_data[test_index, ]
  
  lda_cv_model <- lda(
    Condition ~ .,
    data = train_cv
  )
  
  lda_cv_pred[test_index] <- as.character(
    predict(
      lda_cv_model,
      newdata = test_cv
    )$class
  )
}

## Convert predictions to a factor
lda_cv_pred <- factor(
  lda_cv_pred,
  levels = c("CondA", "CondB")
)

## Calculate cross-validation performance
lda_cv_cm <- confusionMatrix(
  data = lda_cv_pred,
  reference = ML_data$Condition,
  positive = "CondB"
)

lda_cv_cm
```

### Step 9.3: SVM 7-fold Cross-Validation

```{r}
## Create an empty vector - stores SVM predictions
svm_cv_pred <- rep(NA, nrow(ML_data))

## Perform 7-fold cross-validation
for (i in 1:length(folds)) {
  
  test_index <- folds[[i]]
  
  train_cv <- ML_data[-test_index, ]
  test_cv <- ML_data[test_index, ]
  
  svm_cv_model <- svm(
    Condition ~ .,
    data = train_cv,
    kernel = "radial"
  )
  
  svm_cv_pred[test_index] <- as.character(
    predict(
      svm_cv_model,
      newdata = test_cv
    )
  )
}

## Convert predictions to a factor
svm_cv_pred <- factor(
  svm_cv_pred,
  levels = c("CondA", "CondB")
)

## Calculate cross-validation performance
svm_cv_cm <- confusionMatrix(
  data = svm_cv_pred,
  reference = ML_data$Condition,
  positive = "CondB"
)

svm_cv_cm
```

### Step 9.4: Logistic regression 7-fold Cross-Validation

```{r}
## Create an empty vector - stores logistic regression predictions
log_cv_pred <- rep(NA, nrow(ML_data))

## Perform 7-fold cross-validation
for (i in 1:length(folds)) {
  
  test_index <- folds[[i]]
  
  train_cv <- ML_data[-test_index, ]
  test_cv <- ML_data[test_index, ]
  
  log_cv_model <- glm(
    Condition ~ .,
    data = train_cv,
    family = "binomial"
  )
  
  log_cv_prob <- predict(
    log_cv_model,
    newdata = test_cv,
    type = "response"
  )
  
  log_cv_pred[test_index] <- ifelse(
    log_cv_prob >= 0.5,
    "CondB",
    "CondA"
  )
}

## Convert predictions to a factor
log_cv_pred <- factor(
  log_cv_pred,
  levels = c("CondA", "CondB")
)

## Calculate cross-validation performance
log_cv_cm <- confusionMatrix(
  data = log_cv_pred,
  reference = ML_data$Condition,
  positive = "CondB"
)

log_cv_cm

```

### Step 9.5: Comparison of Cross-Validation Classifier Performance

```{r}
## Summarise 7-fold cross-validation performance

cv_performance <- data.frame(
  Model = c(
    "Logistic Regression",
    "LDA",
    "SVM (Radial)"
  ),
  
  Accuracy = c(
    unname(log_cv_cm$overall["Accuracy"]),
    unname(lda_cv_cm$overall["Accuracy"]),
    unname(svm_cv_cm$overall["Accuracy"])
  ),
  
  Sensitivity = c(
    unname(log_cv_cm$byClass["Sensitivity"]),
    unname(lda_cv_cm$byClass["Sensitivity"]),
    unname(svm_cv_cm$byClass["Sensitivity"])
  ),
  
  Specificity = c(
    unname(log_cv_cm$byClass["Specificity"]),
    unname(lda_cv_cm$byClass["Specificity"]),
    unname(svm_cv_cm$byClass["Specificity"])
  ),
  
  Balanced_Accuracy = c(
    unname(log_cv_cm$byClass["Balanced Accuracy"]),
    unname(lda_cv_cm$byClass["Balanced Accuracy"]),
    unname(svm_cv_cm$byClass["Balanced Accuracy"])
  )
)

cv_performance

```

### Step 9.6: Cross-Validation of SVM Kernel Function

Since our earlier single split had radial, polynomial and sigmoid tied. It would be useful to see whether that tie survives 7-fold CV?

```{r}
kernels <- c("radial", "linear", "polynomial", "sigmoid")

## Create an empty dataframe - stores results
svm_kernel_cv <- data.frame()

for (kernel_name in kernels) {
  
  kernel_pred <- rep(NA, nrow(ML_data))
  
  for (i in 1:length(folds)) {
    
    test_index <- folds[[i]]
    
    train_cv <- ML_data[-test_index, ]
    test_cv <- ML_data[test_index, ]
    
    kernel_model <- svm(
      Condition ~ .,
      data = train_cv,
      kernel = kernel_name
    )
    
    kernel_pred[test_index] <- as.character(
      predict(
        kernel_model,
        newdata = test_cv
      )
    )
  }
  
  kernel_pred <- factor(
    kernel_pred,
    levels = c("CondA", "CondB")
  )
  
  kernel_cm <- confusionMatrix(
    data = kernel_pred,
    reference = ML_data$Condition,
    positive = "CondB"
  )

  svm_kernel_cv <- rbind(
    svm_kernel_cv,
    data.frame(
      Kernel = kernel_name,
      Accuracy = unname(kernel_cm$overall["Accuracy"]),
      Sensitivity = unname(kernel_cm$byClass["Sensitivity"]),
      Specificity = unname(kernel_cm$byClass["Specificity"]),
      Balanced_Accuracy =
        unname(kernel_cm$byClass["Balanced Accuracy"])
    )
  )
}


svm_kernel_cv
```

## Step 10: Results:

Logistic regression: **54% CV accuracy**

LDA: **74% CV accuracy**

Radial SVM: **84% CV accuracy**

Best SVM configuration — sigmoid: **88% CV accuracy**

### Step 10.1: Visualisation of results

```{r}
## Build classifier performance table from saved results

classifier_plot_data <- data.frame(
  Model = c(
    "Logistic Regression",
    "LDA",
    "SVM (Sigmoid)"
  ),

  Accuracy = c(
    unname(log_cv_cm$overall["Accuracy"]),
    unname(lda_cv_cm$overall["Accuracy"]),
    svm_kernel_cv$Accuracy[svm_kernel_cv$Kernel == "sigmoid"]
  ),

  Sensitivity = c(
    unname(log_cv_cm$byClass["Sensitivity"]),
    unname(lda_cv_cm$byClass["Sensitivity"]),
    svm_kernel_cv$Sensitivity[svm_kernel_cv$Kernel == "sigmoid"]
  ),

  Specificity = c(
    unname(log_cv_cm$byClass["Specificity"]),
    unname(lda_cv_cm$byClass["Specificity"]),
    svm_kernel_cv$Specificity[svm_kernel_cv$Kernel == "sigmoid"]
  )
)

classifier_plot_data


##Reshape
classifier_long <- reshape2::melt(
  classifier_plot_data,
  id.vars = "Model",
  variable.name = "Metric",
  value.name = "Performance"
)


## Plot
#| fig-width: 9
#| fig-height: 6
#| out-width: "100%"

ggplot(
  classifier_long,
  aes(
    x = Model,
    y = Performance,
    fill = Metric
  )
) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7
  ) +

  geom_text(
    aes(
      label = paste0(
        round(Performance * 100),
        "%"
      )
    ),
    position = position_dodge(width = 0.8),
    vjust = -0.4,
    size = 3.5
  ) +

  scale_y_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    labels = scales::percent_format()
  ) +

  labs(
    title = "Seven-fold cross-validated classifier performance",
    x = NULL,
    y = "Performance",
    fill = "Metric"
  ) +

  theme_classic(base_size = 12) +

  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    ),

    axis.text.x = element_text(
      angle = 20,
      hjust = 1
    ),

    legend.position = "top"
  )
```
Packages used:
```{r}
sessionInfo()
```
